Factors affecting inferences on natural mortality and associated environmental effects in state-space age-structured assessment models

### 2025-05-29 Ecov_study/mortality/manuscript/manuscript.Rmd
saving work on ecov-m

All of our analyses used the Woods Hole Assessment Model (WHAM) to construct both OMs and EMs [@millerstock20; @stockmiller21; @milleretal_inreview]. The WHAM package has been used extensively to configure OMs and EMs for several other simulation studies [@legaultetal23; @lietal24; @brittenetal_inreview; @lietal_inreview_a] and is used to assess many commercially important stocks in the Northeast U.S. [e.g., @nefsc22;@nefsc22a;@nefsc24]. We used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3) for to generate all results.

@@

In the WHAM model, environmental covariates are assumed to be described as state-space processes with annual observations of the true latent covariate [@milleretal16;stockmiller21]. In our simulations, the latent covariate is assumed to be a stationary first order autoregressive (AR1) process
$$
X_y|X_{y-1} \sim \text{N}\left(\mu_E\left(1-\rho_E\right) + \rho_E E_{y-1}, \left(1-\rho_E^2\right)\sigma^2_E\right)
$$
with marginal mean $\mu_E=0$ and variance $\sigma^2_E$. The four configuration of the latent environmental covariate in the operating models assume the one of two values for the marginal standard deviation ($\sigma_E \in \{0.1, 0.5\}$) and for the autocorrelation parameter ($\rho_E \in \{0, 0.5\}$).

The observations of the latent environmental covariate are assumed to be unbiased and Gaussian
$$
x_y|X_y \sim \text{N}\left(X_y,\sigma^2_e\right)
$$
The standard deviation of the environmental observations in the operating models is one of two values $\sigma_E \in \{0.1, 0.5\}$. Figure \ref{om_ecov_example} provides example simulations of the latent process and observations under the alternative configurations.

@@

Many of the characteristics of the population biology and structure are the same as those in @milleretal_inreview1. We configures the stock with 10 age classes (ages 1 to 10+) where the last is a "plus" group that accumulates any older individuals. We model the population for 40 years with annual catch and index observations. We assume spawning occurs annually 1/4 of the way through the year. The maturity at age was a logistic function of age with age at 50\% maturity ($a_{50}$) = 2.89 and slope = 0.88 (Figure \ref{om_inputs_fig}, top left). We assumed a von Beralanffy growth function for weight at age  with length at age defined as 
$$
L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right)
$$
where $t_0 = 0$, $L_\infty = 85$, and $k = 0.3$, and a length-weight 
$$
W_a = \theta_1 L_a^{\theta_2}
$$
where $\theta_1 = e^{-12.1}$ and $\theta_2 = 3.2$ (Figure \ref{om_inputs_fig}, top right).

The general model for natural mortality in year $y$ is a log-linear function of a mean(median) parameter $\beta_{M}$, effects of an environmental covariate $E_y$ and a possibly autocorrelated normally distributed annual process error $\varepsilon_{M,y}$
$$
\log M_y = \beta_M + \beta_{E} E_y + \varepsilon_{M,y}
$$
$\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)$ [@stockmiller21]. We assume the median natural mortality rate $e^{\beta_M} = 0.2$ is constant across ages. For R and R+S OMs and EMs, the process errors are set equal to 0. For all R+M OMs, we assume the same standard deviation $\sigma_M = 0.3$ and is estimated in the R+M EMs. The covariate effect is one of 3 alternative values in the operating models ($\beta_\text{E} \in \{0,0.25,0.5\}$. The parameters defining the simulated covariate time series, size of the covariate effect, and any natural mortality random effects result in a range of different levels of variation in annual natural mortality rates (Figure \ref{M_example}).

We assume is a single fishing fleet harvesting the population with a logistic selectivity function with age at 50\% selection $a_{50} = 5$ and slope = 1 (Figure \ref{om_inputs_fig}, bottom left).

We assumed expected recruitment each year from a Beverton-Holt stock-recruit relationship (SRR)
\[
R_{y} = \frac{a \text{SSB}_{y-1}}{1 + b \text{SSB}_{y-1}}.
\]
All biological inputs to calculations of spawning biomass per recruit (i.e., weight, maturity, and natural mortality at age) are constant in the R and R+S OMs without covariate effects on natural mortality. Therefore, steepness and unfished recruitment are also constant over the time period for those OMs [@millerbrooks21] and with our assumed biological inputs and selectivity the fishing mortality resulting in 40\% reduction in spawning biomass per recruit is $F_{40\%} = 0.348$. With an assumed unfished recruitment of $R_0 = e^{10}$, setting $\Fmsy = \Fspr[40]$ results in a steepness of 0.69 and $a=0.60$ and $b = 2.4 \times 10^{-5}$ (Figure \ref{om_inputs_fig}, bottom right).

We used two fishing scenarios for OMs. In the first scenario, the stock experiences overfishing at 2.5\Fmsy for the first 20 years followed by fishing at \Fmsy for the last 20 years (denoted $2.5\Fmsy \rightarrow \Fmsy$). In the second scenario, the stock is fished at \Fmsy for the entire time period (40 years). The magnitude of the overfishing assumptions is based on average estimates of overfishing for NEUS groundfish stocks from @wiedenmannetal19 and similar to the approach in @legaultetal23.

We specified initial population abundance at age at the equilibrium distribution that corresponds to fishing at either $F = 2.5\times \Fmsy$ or $F = \Fmsy$. This implies that, for a deterministic model, the abundance at age would not change from the first year to the next.

For R+M OMs and all OMs with environmental covariate effects on natural mortality, steepness is not constant, but we used the same $a$ and $b$ parameters as other operating models which equates to a steepness and $R_0$ at the median of the time series process for M. For operating models with time-varying random effects for fishery selectivity, \Fmsy is also not constant however we use the same fishing history as other operating models which corresponds to \Fmsy at the mean selectivity parameters.

@@

We assumed a single fleet operating year round for catch observations with logistic selectivity for the fleet with $a_{50} = 5$ and slope = 1 (Figure \ref{om_inputs_fig}). This selectivity was used to define \Fmsy, \Fspr[40], and steepness for the Beverton-Holt stock recruitment parameters above. We assumed a logistic-normal distribution for the age-composition observations associated with the fleet where errors in the multivariate normal transformation are independent. The standard deviation parameter is also constant across ages.

Two time series of surveys are assumed and observed in numbers rather than biomass for the entire 40 year period with one occurring in the spring (0.25 way through the year) and one in the fall (0.75 way through the year). Catchability of both surveys are assumed to be 0.1.  We assumed the same selectivity and age composition distribution as the fleet for both indices.

@@

For each data set simulated from an operating model 12 estimating models were fit. There were three factors defining the configuration of each estimating model

\begin{itemize}
\item whether the mean natural mortality $\beta_M$ was estimated or assumed known ($\log 0.2$)
\item whether an environmental effect $\beta_E$ was estimated or not (fixed at 0)
\item whether the process errors were assumed on recruitment only (R), recruitment and survival (R+S), or recruitment and natural mortality (R+M)
\end{itemize}
The configuration of the process errors in the estimating models generally matched the corresponding options in the operating models.

For example, uncorrelated R+S was assumed for both the estimating and operating model. However, R+M EMs did not assume M random effects were uncorrelated (parameter was estimated). The environmental covariate observations were included in all estimation models to ensure comparability of information criteria. All fixed effects parameters for selectivity, catchability, fully-selected fishing mortality, mean recruitment, initial abundance at age, and variances for logistic-normal age composition distributions were estimated. Any process error variance parameters for recruitment, survival, and natural mortality were also estimated. The observation error variance of the environmental observations and aggregate catch and indices were all assumed known at the true values.

@@

### Convergence {-}

@@

We measured the frequency of correct model selection using marginal AIC. For a given operating model the set of models that were considered all made the same assumptions on whether or not to estimate (mean) natural mortality rate or it is assumed at the true value $\beta_M = log(0.2)$. For model $m$, the marginal AIC is a function of the marginal log-likelihood maximized with respect to the fixed effects in the model $\boldsymbol{\theta}$ and the number of fixed effects $n\left(\boldsymbol{\theta}\right)$ estimated,
$$
\text{AIC}_m = -2\left[{\text{argmax}}_{\boldsymbol{\theta}} \log L_m\left({\boldsymbol{\theta}}\right) - n\left({\boldsymbol{\theta}}\right)\right].
$$

### Bias {-}

We calculated median errors of
\begin{itemize}
\item $\beta_E$, the effect of environmental covariates on natural mortality, 
\item $\beta_M$, the mean log-natural mortality rate,
\end{itemize}
and the median relative errors of 
\begin{itemize}
\item annual natural mortality rate,
\item annual spawning stock biomass, and
\item annual fully-selected fishing mortality rate.
\end{itemize}

Note that the exponentiating the median error of $\beta_M$ would be equivalent to the median relative error of the median natural mortality rate because the median of the exponential is the same as the exponential of the median.

Results for fishing mortality rate are provided in the Supplementary Materials. For the $i$th simulated data the relative error for a parameter $\theta$ provided from the fitted estimation model is
$$
{\text{RE}_i}\left(\theta\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
$$
and measured bias as the median relative error and constructed 95\% confidence intervals using the binomial distribution approach as in @stockmiller21. If the confidence interval contains zero we do not have evidence of bias from the simulation study. We used estimates from all simulations that satisfied the first type of convergence described above. We were not concerned about more restrictive types of convergence because some estimation models would not be expected to satisfy these criteria because of mis-specified structures. For example, an estimation model that includes survival random effects fitted to a set of observations simulated without survival random effects should have trouble estimating variance of the survival random effects, but the estimates of the parameters and derived output described above, might be reliable.

For natural mortality rate, SSB, and fully-selected fishing mortality rate, we summarized results for each of the annual values, but we present results only for estimates from the first year (start), year 21 (middle), and the last year (end), because there were no appreciable differences between the results for these three years and those for the other years.

@@

## Convergence {-}

@@

@@

When the EMs assumed the mean natural mortality rate was known, we observed generally accurate estimation of environmental effects across all EM and OM process error assumptions and all true covariate effect sizes, when ther was low uncertainty in environmental observations and larger temporal contrast in the simulated true environmental covariate (Figure \ref{Ecov_beta_bias_M_fixed}). We observed a negative trend in bias of the environmental effect with increased effect size when temporal variation in the covariate was lower and/or uncertainty in the covariate observations was higher. When the OMs had R+S process errors with low temporal variation in the true environmental covariate and lower uncertainty in the indices age age composition, estimated covariate effects were highly variable.  In most cases the relative error of $\beta_E$ did not depend on the source of process error assumed in the EM. When there was an effect of the EM process error assumption it was when OMs had R+S process errors. The worst bias was observed when OMs assumed R+S process errors, high uncertainty in covariate observations, low variability in the covariate, and low uncertainty in index and age composition observations.

When the mean natural mortality rate was estimated in the EM, results were similar except estimated effects were even more variable for data simulated with R+S process errors (Figure \ref{Ecov_beta_bias_M_estimated}). There was also more separation of reliability of the estimation of the effect among EMs with different process error assumptions. The separation was most apparent when OMs simulated R+S process errors and larger variability in the environmental covariate wehere EMs with process errors other than R+S showing more bias than when the mean natural mortality rate was assumed known.

### Mean natural mortality rate {-}

We found high accuracy for estimation of the mean natural mortality rate parameter ($\beta_M$) for all EM process errors assumptions when OMs had step changes in fishing mortality, lower uncertainty in index and age composition observations, and either R or R+M process errors (Figure \ref{mean_M_bias_Ecov_beta_fixed}). The most variation in estimates occurred when fishing mortality was constant and there was higher uncertainty in index and age composition observations. For OMs with R+S process errors, the most reliable estimation of $\beta_M$ was obtained when the EM also assumed R+S process errors across all other factors. For OMs with R+M process errors, the matching assumption for the EM only showed the best reliability when there was low uncertainty in index and age composition observations and fishing mortality was constant.

@@

When OMs and EMs assume $\beta_M = log(0.2)$ and $\beta_E = 0$, there is no annual variation in natural mortality for OMs with R or R+S process errors simulated or assumed in the EM and, therefore, estimation bias is not possible (Figure \ref{annual_M_bias_M_fixed_beta_fixed}). We also observe little or no evidence of bias (confidence intervals include 0) for R or R+S EMs for any OMs even when the OMs included an effect of the covariate on natural mortality ($\beta_E > 0$). Including process error on M (R+M) produces more variability in errors for R and R+S EMs than including increased level of effect of the covariate on M. For R and R+S OMs, errors in annual M for R+S EMs were less variable than those for R EMs.

@@

When $\beta_E$ was estimated, R EMs performed worse for R+S OMs with evidence of negative bias for OMs with constant fishing mortality, lower uncertainty in index and age composition observations, the largest covariate effect size, lower variability in the latent covariate, and larger uncertainty in the covariate observations (Figure \ref{annual_M_bias_M_fixed_beta_estimated}). There was little difference in the results for R+S EMs whether $\beta_E$ was estimated or not. The results for R+M EMs were generally similar to those when $\beta_E=0$ was assumed, except that the differences between the estimates at the beginning of the time series and later years for certain OM configurations did not occur.

Allowing the EMs to estimate $\beta_M$ resulted in very large variability in estimates for all EMs for R and R+M OMs with constant fishing mortality and higher uncertainty in index and age composition observations (Figure \ref{annual_M_bias_M_estimated_beta_estimated}). The same variability occurred for R+S OMs, but strong bias was estimated for R and R+M EMs. Annual M estimation was most reliable when OMs had step changes in fishing mortality and lower uncertainty in index and age composition observations. For R and R+M OMs with those characteristics, all EMs generally provided accurate estimation of annual M, but only R+S EMs provided accurate estimation for R+S OMs. Little or no bias in annual M estimation was observed for R+S EMs across all OM process error assumptions as long as there was a step change in fishing mortality and lower uncertainty in indices and age composition.

@@

When EMs assume $\beta_M = log(0.2)$ and $\beta_E = 0$, there is little evidence of bias for R and R+M OMs except when there is a step change in fishing mortality and higher uncertainty in index and age composition observations, particularly at the end of the time series (Figure \ref{SSB_bias_M_fixed_beta_fixed}). We observed a similar trend in median relative error for R+S OMs, but there was more evidence of positive bias at the beginning of the time series whereas the confidence intervals for the negative median errors at the end of the time series often included 0. However, there was more indication of positive bias of the incorrect process error assumption of the EMs for the R+S OMs. For R+S EMs fit to R+S OMs, there was indication of small positive bias at the beginning of the time series when there was constant fishing mortality and higher uncertainty in index and age composition observations.

When EMs assume $\beta_M = log(0.2)$, but estimate $\beta_E$, the median errors for SSB for R and and R+M OMs are similar to those when $\beta_E = 0$ (Figure \ref{SSB_bias_M_fixed_beta_estimated}). For R+S OMs, the median errors for EMs that also assume R+S process errors are also similar to those with no covariate effect assumed. However, for R and R+M EMs fit to R+S OMs with constant fishing mortality rate and lower uncertainty in indices and age composition observations, we observed evidences of negative bias when higher covariate observation uncertainty and lower variation in the latent covariate was simulated and positive bias for other configurations of the covariate and corresponding observation uncertainty. When R+S OMs had constant fishing mortality rate and higher uncertainty in indices and age composition observations or step changes in fishing mortality and lower uncertainty in indices and age composition, R and R+M EMs often provided positively biased SSB estimates when there was lower uncertainty in covariate observations.

When EMs estimated both $\beta_M$ and $\beta_E$, we observed large variation in the errors in SSB under the same OM configurations where we observed large variation in errors for annual natural mortality rates (Figure \ref{SSB_bias_M_estimated_beta_estimated}). Similarly, we observed little or no bias in SSB estimation for R+S EMs across all OM process error assumptions as long as there was a step change in fishing mortality and lower uncertainty in indices and age composition.

@@

@@

@@

\begin{landscape}
\begin{figure}
\begin{center}
\includegraphics{convergence}
\end{center}
\caption{Estimated probability of fits providing hessian-based standard errors for EMs assuming alternative process error, that estimate or assume known median natural mortality, and that estimate or assume no covariate effect on median natural mortality when fitted to operating models that have R (left) and R+S (middle) , or R+M (right) process error structures and three levels of true covariate effect on median natural mortality (x axis). Vertical lines represent 95\% confidence intervals.}\label{convergence}
\end{figure}
\end{landscape}

\begin{landscape}
\begin{figure}
\begin{center}
\includegraphics[height = \textheight]{aic_main}
\end{center}
\caption{Proportion of simulated data sets for each operating model where the estimation model type (treatment of environmental covariate effect and  assumed process error type) had the lowest AIC. All OMs had low observation error for fish population observations and temporal contrast in fishing pressure. All EMs estimated median natural mortality rate.}\label{aic}
\end{figure}
\end{landscape}

\begin{landscape}
\begin{figure}
\begin{center}
%\includegraphics[height = \textheight]{Ecov_beta_bias}
\end{center}
\caption{Relative error of estimates of environmental effect on natural mortality $\beta_E$ from fitting EMs with alternative process error assumptions. All OMs had low observation error and contrast in fishing mortality. Vertical lines represent 95\% confidence intervals.}\label{ecov_beta_bias}
\end{figure}
\end{landscape}

\begin{landscape}
\begin{figure}
\begin{center}
%\includegraphics[height = \textheight]{median_log_M_bias}
\end{center}
\caption{Relative error of the estimated mean log-natural mortality parameter $\beta_{M}$ from fitting EMs with alternative process error assumptions. All OMs had low observation error and contrast in fishing mortality. Vertical lines represent 95\% confidence intervals.}\label{median_log_M_bias}
\end{figure}
\end{landscape}

>>>>

All of our analyses used the Woods Hole Assessment Model (WHAM) to construct both OMs and EMs [@millerstock20; @stockmiller21; @milleretal_inreview]. The WHAM package has been used extensively to configure OMs and EMs for several other simulation studies [@stocketal21;@legaultetal23; @lietal24; @brittenetal_inreview; @lietal_inreview_a] and is used to assess many commercially important stocks in the Northeast U.S. [e.g., @nefsc22;@nefsc22a;@nefsc24]. We used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3) for to generate all results.

@@

In the WHAM model, environmental covariates are assumed to be described as state-space processes with annual observations of the true latent covariate [@milleretal16;@stockmiller21]. In our simulations, the latent covariate is assumed to be a stationary first order autoregressive (AR1) process
$$
X_y|X_{y-1} \sim \text{N}\left(\mu_E\left(1-\rho_E\right) + \rho_E X_{y-1}, \left(1-\rho_E^2\right)\sigma^2_E\right)
$$
with marginal mean $\mu_E=0$ and variance $\sigma^2_E$. The four configuration of the latent environmental covariate in the operating models assume the one of two values for the marginal standard deviation $\sigma_E \in \{0.1, 0.5\}$ and for the autocorrelation parameter $\rho_E \in \{0, 0.5\}$.

The observations of the latent environmental covariate are assumed to be unbiased and Gaussian
$$
x_y|X_y \sim \text{N}\left(X_y,\sigma^2_e\right)
$$
The standard deviation of the environmental observations in the operating models is one of two values $\sigma_e \in \{0.1, 0.5\}$. Figure \ref{om_ecov_example} provides example simulations of the latent process and observations under the alternative configurations.

@@

Many of the characteristics of the population biology and structure including the age classes (10 age classes (ages 1 to 10+)), time span (40 years), maturity (Figure \ref{om_inputs_fig}, top left), growth  (Figure \ref{om_inputs_fig}, top right), time of spawning (1/4 of the year), and recruitment (Figure \ref{om_inputs_fig}, bottom right) are identical to @milleretal_inreview1. The maturity at age is a logistic function of age with age at 50\% maturity ($a_{50}$) = 2.89 and slope = 0.88 and weight at age is derived from a von Beralanffy growth function where $t_0 = 0$, $L_\infty = 85$, and $k = 0.3$, and a length-weight relationship
$$
W_a = \theta_1 L_a^{\theta_2}
$$
where $\theta_1 = e^{-12.1}$ and $\theta_2 = 3.2$.

Here we consider both covariate effects and process errors for natural mortality and the general model for natural mortality in year $y$ is a log-linear function of a  parameter $\beta_{M}$ that defines median natural mortality and possibly effects of the environmental covariate $X_y$ described above and autocorrelated normally distributed annual process error $\varepsilon_{M,y}$
$$
\log M_y = \beta_M + \beta_{E} X_y + \varepsilon_{M,y}
$$
$\epsilon_{M,y}|\epsilon_{M,y-1} \sim \text{N}\left(\epsilon_{M,y-1},(1-\rho_M^2)\sigma_M^2\right)$ [@stockmiller21]. We assume the median natural mortality rate $e^{\beta_M} = 0.2$ is constant across ages. For R and R+S OMs and EMs, $\epsilon_{M,y} = 0$. For all R+M OMs, we assume the same standard deviation $\sigma_M = 0.3$ and is estimated in the R+M EMs. The covariate effect is one of 3 alternative values in the operating models ($\beta_\text{E} \in \{0,0.25,0.5\}$. The parameters defining the simulated covariate time series, size of the covariate effect, and any natural mortality random effects result in a range of different levels of variation in annual natural mortality rates (Figure \ref{M_example}).

We assumed expected recruitment each year from a Beverton-Holt stock-recruit relationship (SRR)
\[
R_{y} = \frac{a \text{SSB}_{y-1}}{1 + b \text{SSB}_{y-1}}.
\]
All biological inputs to calculations of spawning biomass per recruit (i.e., weight, maturity, and natural mortality at age) are constant in the R and R+S OMs without covariate effects on natural mortality. Therefore, steepness and equilibrium unfished recruitment are also constant over the time period for those OMs [@millerbrooks21]. As in @milleretal_inreview1, our assumed biological inputs and selectivity (defined below) with constant natural mortality result in equilibrium fishing mortality that reduces spawning biomass per recruit to 40\% of the unfished level is $\Fspr[40] = 0.348$. With an assumed unfished recruitment of $R_0 = e^{10}$, setting $\Fmsy = \Fspr[40]$ results in a steepness of 0.69 and $a=0.60$ and $b = 2.4 \times 10^{-5}$.

For R+M OMs and all OMs with covariate effects on natural mortality, steepness is not constant, but we used the same $a$ and $b$ parameters as other operating models which equates to a steepness and $R_0$ at the median of the time series process for M. For operating models with time-varying random effects for fishery selectivity, \Fmsy is also not constant however we use the same fishing history as other operating models which corresponds to \Fmsy at the mean selectivity parameters.

We also used the same two fishing scenarios as @milleretal_inreview1 for OMs. In the first scenario, the stock experiences overfishing at 2.5\Fmsy for the first 20 years followed by fishing at \Fmsy for the last 20 years (denoted $2.5\Fmsy \rightarrow \Fmsy$). In the second scenario, the stock is fished at \Fmsy for the entire time period (40 years).

@@

We define the generation of observations of total catch, aggregate indices, and corresponding age composition identical to @milleretal_inreview1. There is a single fleet operating year round for catch observations with logistic selectivity for the fleet with $a_{50} = 5$ and slope = 1 (Figure \ref{om_inputs_fig}, bottom left). Observations are generated for all 40 years of the model. There are two index time series intended to represent fishery-independent surveys occurring in the spring (0.25 way through the year) and the fall (0.75 way through the year). Catchability of both surveys are assumed to be 0.1. We assumed catch and index age composition observations aer generated from a logistic-normal distribution where errors on the multivariate normal scale are independent. The standard deviation parameter is also constant across ages.

@@

For each data set simulated from an operating model 12 estimating models were fit. There were three factors defining the configuration of each estimating model: 1) whether the mean natural mortality $\beta_M$ was estimated or assumed known, 2) whether an environmental effect $\beta_E$ was estimated or not, and 3) whether the process errors were assumed on recruitment only (R), recruitment and survival (R+S), or recruitment and natural mortality (R+M).

The configuration of the process errors in the estimating models generally matched the corresponding options in the operating models. For example, uncorrelated R+S was assumed for both the estimating and operating model. However, R+M EMs did not assume M random effects were uncorrelated ($\rho_M$ was estimated). The environmental covariate observations were included in all estimation models to ensure comparability of AIC. All fixed effects parameters for selectivity, catchability, fully-selected fishing mortality, mean recruitment, initial abundance at age, and variances for logistic-normal age composition distributions were estimated. Any process error variance parameters for recruitment, survival, and natural mortality were also estimated. The observation error variance of the environmental observations and aggregate catch and indices were all assumed known at the true values.

@@

### EM convergence {-}

@@

We measured the frequency of correct model selection using marginal AIC. For a given operating model the set of models that were considered all made the same assumptions on whether or not to estimate (mean) natural mortality rate or it is assumed at the true value $\beta_M = log(0.2)$. For model $m$, the marginal AIC is a function of the marginal log-likelihood maximized with respect to the fixed effects in the model $\boldsymbol{\theta}$ and the number of fixed effects $n\left(\boldsymbol{\theta}\right)$ estimated,
$$
\text{AIC}_m = -2\left[{\text{argmax}}_{\boldsymbol{\theta}} \log L_m\left({\boldsymbol{\theta}}\right) - n\left({\boldsymbol{\theta}}\right)\right].
$$
All model fits that successfully completed the optimization were used in these results. We used all of these fits because some lack of convergence would be expected for the correct behavior of more complicated models that include process errors that did not exist in the operating model. For example R+M EMs fit to R OMs would be expected to estimate no variance in the natural mortality  random effects and the estimated variance parameter going to zero would cause poor convergence.

### Parameter inference reliability {-}

All results here use OM simulations with fits that satisfied the convergence criterion described above. We used this conditioning, to reflect how practitioners would proceed in analyses of model fits with real assessment data. That is, practitioners would ensure models converged such that hessian-based standard errors were available for all model parameter estimates.

We calculated median errors (ME) of $\beta_E$, the effect of environmental covariates on natural mortality, and the median natural mortality rate parameter $\beta_M$ (the mean log-natural mortality rate). We calculated the median relative errors (MRE) of terminal year estimates of natural mortality rate, spawning stock biomass, and fully-selected fishing mortality rate.
In preliminary analyses we inspected MRE of these parameters over the whole time series, but we found no appreciable differences in patterns across the various factors defining the OMs and EMs. Furthermore, results for terminal year fishing mortality were generally inversely related to those for spawning stock biomass, and are provided in the Supplementary Materials and not discussed further.

We also calculated the root mean square error (RMSE) and estimated probability of coverage of constructed 95\% confidence intervals for $\beta_E$ and $\beta_M$ for EMs that estimated these parameters. Confidence intervals for probabilities of coverage are constructed using the same methods as those for probabilities of convergence.

Results for fishing mortality rate For the $i$th simulated data the relative error for a parameter $\theta$ provided from the fitted estimation model is
$$
{\text{RE}_i}\left(\theta\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
$$
and  constructed 95\% confidence intervals using the binomial distribution approach as in @stockmiller21 and @milleretal_inreview1. Because of the inverse relationship of the the confidence interval and the null hypothesis significance test, when the confidence interval contains zero we do not have evidence of bias from the simulation study.

For natural mortality rate, SSB, and fully-selected fishing mortality rate, we summarized results for each of the annual values, but we present results only for estimates from the terminal year, because there were no appreciable differences between the results among all years.

@@

## EM Convergence {-}

@@

We only present results for EMs where the median natural mortality rate parameter was estimated. Results differed little between results where this parameter was estimated or assumed known, but all of these results can be found in the Supplementary materials.

@@

When the EMs assumed the median natural mortality rate was known, we observed generally accurate estimation of environmental effects across all EM and OM process error assumptions and all true covariate effect sizes, when ther was low uncertainty in environmental observations and larger temporal contrast in the simulated true environmental covariate (Figure \ref{beta_E_bias}). We observed a negative trend in bias of the environmental effect with increased effect size when temporal variation in the covariate was lower and/or uncertainty in the covariate observations was higher. When the OMs had R+S process errors with low temporal variation in the true environmental covariate and lower uncertainty in the indices age age composition, estimated covariate effects were highly variable.  In most cases the relative error of $\beta_E$ did not depend on the source of process error assumed in the EM. When there was an effect of the EM process error assumption it was when OMs had R+S process errors. The worst bias was observed when OMs assumed R+S process errors, high uncertainty in covariate observations, low variability in the covariate, and low uncertainty in index and age composition observations.

When the median natural mortality rate was estimated in the EM, results were similar except estimated effects were even more variable for data simulated with R+S process errors (Figure \ref{beta_E_bias}). There was also more separation of reliability of the estimation of the effect among EMs with different process error assumptions. The separation was most apparent when OMs simulated R+S process errors and larger variability in the environmental covariate wehere EMs with process errors other than R+S showing more bias than when the mean natural mortality rate was assumed known.

### Median natural mortality rate {-}

We found high accuracy for estimation of the median natural mortality rate parameter ($\beta_M$) for all EM process errors assumptions when OMs had step changes in fishing mortality, lower uncertainty in index and age composition observations, and either R or R+M process errors (Figure \ref{beta_M_bias}). The most variation in estimates occurred when fishing mortality was constant and there was higher uncertainty in index and age composition observations. For OMs with R+S process errors, the most reliable estimation of $\beta_M$ was obtained when the EM also assumed R+S process errors across all other factors. For OMs with R+M process errors, the matching assumption for the EM only showed the best reliability when there was low uncertainty in index and age composition observations and fishing mortality was constant.

@@

When OMs and EMs assume $\beta_M = log(0.2)$ and $\beta_E = 0$, there is no annual variation in natural mortality for OMs with R or R+S process errors simulated or assumed in the EM and, therefore, estimation bias is not possible (Figure \ref{terminal_M_bias}). We also observe little or no evidence of bias (confidence intervals include 0) for R or R+S EMs for any OMs even when the OMs included an effect of the covariate on natural mortality ($\beta_E > 0$). Including process error on M (R+M) produces more variability in errors for R and R+S EMs than including increased level of effect of the covariate on M. For R and R+S OMs, errors in annual M for R+S EMs were less variable than those for R EMs.

@@

When $\beta_E$ was estimated, R EMs performed worse for R+S OMs with evidence of negative bias for OMs with constant fishing mortality, lower uncertainty in index and age composition observations, the largest covariate effect size, lower variability in the latent covariate, and larger uncertainty in the covariate observations (Figure \ref{terminal_M_bias}). There was little difference in the results for R+S EMs whether $\beta_E$ was estimated or not. The results for R+M EMs were generally similar to those when $\beta_E=0$ was assumed, except that the differences between the estimates at the beginning of the time series and later years for certain OM configurations did not occur.

Allowing the EMs to estimate $\beta_M$ resulted in very large variability in estimates for all EMs for R and R+M OMs with constant fishing mortality and higher uncertainty in index and age composition observations (Figure \ref{terminal_M_bias}). The same variability occurred for R+S OMs, but strong bias was estimated for R and R+M EMs. Annual M estimation was most reliable when OMs had step changes in fishing mortality and lower uncertainty in index and age composition observations. For R and R+M OMs with those characteristics, all EMs generally provided accurate estimation of annual M, but only R+S EMs provided accurate estimation for R+S OMs. Little or no bias in annual M estimation was observed for R+S EMs across all OM process error assumptions as long as there was a step change in fishing mortality and lower uncertainty in indices and age composition.

@@

When EMs assume $\beta_M = log(0.2)$ and $\beta_E = 0$, there is little evidence of bias for R and R+M OMs except when there is a step change in fishing mortality and higher uncertainty in index and age composition observations, particularly at the end of the time series (Figure \ref{terminal_ssb_bias}). We observed a similar trend in median relative error for R+S OMs, but there was more evidence of positive bias at the beginning of the time series whereas the confidence intervals for the negative median errors at the end of the time series often included 0. However, there was more indication of positive bias of the incorrect process error assumption of the EMs for the R+S OMs. For R+S EMs fit to R+S OMs, there was indication of small positive bias at the beginning of the time series when there was constant fishing mortality and higher uncertainty in index and age composition observations.

When EMs assume $\beta_M = log(0.2)$, but estimate $\beta_E$, the median errors for SSB for R and and R+M OMs are similar to those when $\beta_E = 0$ (Figure \ref{terminal_ssb_bias}). For R+S OMs, the median errors for EMs that also assume R+S process errors are also similar to those with no covariate effect assumed. However, for R and R+M EMs fit to R+S OMs with constant fishing mortality rate and lower uncertainty in indices and age composition observations, we observed evidences of negative bias when higher covariate observation uncertainty and lower variation in the latent covariate was simulated and positive bias for other configurations of the covariate and corresponding observation uncertainty. When R+S OMs had constant fishing mortality rate and higher uncertainty in indices and age composition observations or step changes in fishing mortality and lower uncertainty in indices and age composition, R and R+M EMs often provided positively biased SSB estimates when there was lower uncertainty in covariate observations.

When EMs estimated both $\beta_M$ and $\beta_E$, we observed large variation in the errors in SSB under the same OM configurations where we observed large variation in errors for annual natural mortality rates (Figure \ref{terminal_ssb_bias}). Similarly, we observed little or no bias in SSB estimation for R+S EMs across all OM process error assumptions as long as there was a step change in fishing mortality and lower uncertainty in indices and age composition.

@@

We found estimation of environmental effects on M was possible and reliable in certain scenarios even when the process error was mis-specified (R+S EM and R+M OM) also R EMs were not reliable for Ecov effects when PE was mis-specified. Given that milleretal in review found AIC was accurate for R+S OMs and applications of wham to real assessment data typically finds R+S preferrable to R only, it suggests inferences on covariate effects on M without just R process errors (e.g., deriso et al 2008) should be reevaluated.

@@

We conditioned our investigations of reliability of parameter estimation on simulations with fits that converged satisfactorily. However, these simulations could exclude many realizations of covariate, population, and observation time series that are not sufficiently informative for all the parameters that we were estimating. Why might this matter?

@@

\begin{landscape}
\begin{figure}
\begin{center}
\includegraphics{convergence_main}
\end{center}
\caption{Estimated probability of fits providing hessian-based standard errors for EMs with alternative process error assumptions, treatment of median natural mortality ($e^\beta_M$ known or estimated), and treatment of covariate effect ($\beta_E = 0$ or estimated). The OMs have R (left) and R+S (middle), or R+M (right) process error structures, alternative configurations of covariate time series structure and levels of observation uncertainty (rows), and three levels of true covariate effect on median natural mortality (x axis). All OMs had low observation error for fish population observations and temporal contrast in fishing pressure. Vertical lines represent 95\% confidence intervals.}\label{convergence}
\end{figure}
\end{landscape}

\begin{landscape}
\begin{figure}
\begin{center}
\includegraphics[height = \textheight]{aic_main}
\end{center}
\caption{For each OM, the proportion of simulated data sets where the EM type (treatment of environmental covariate effect and  assumed process error type) had the lowest AIC. All OMs had low observation error for fish population observations and temporal contrast in fishing pressure. All EMs estimated median natural mortality rate.}\label{aic}
\end{figure}
\end{landscape}

\begin{landscape}
\begin{figure}
\begin{center}
\includegraphics[height = \textheight]{beta_E_bias_main}
\end{center}
\caption{Median error (ME) of estimates of environmental effect on natural mortality $\beta_E$ from fitting EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated). All OMs had low observation error and contrast in fishing mortality. Vertical lines represent 95\% confidence intervals.}\label{beta_E_bias}
\end{figure}
\end{landscape}

\begin{landscape}
\begin{figure}
\begin{center}
\includegraphics[height = \textheight]{beta_E_CI_coverage_main}
\end{center}
\caption{Probability of 95\% confidence interval for $\beta_E$ containing the true value for EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated). All OMs had low observation error and contrast in fishing mortality. Vertical lines represent 95\% confidence intervals.}\label{beta_E_CI_coverage}
\end{figure}
\end{landscape}
